Equalizer\\Programming and User Guide

### 2011-08-10 Developer/ProgrammingGuide/paper.tex
Programming Guide: add focus distance section
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8.2 for Equalizer 1.0+git \\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.8.1   & Mar 25, 2011 & Minor Revision\\
    1.8.2   & Apr 04, 2011 & Added \sref{sFocusDistance}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\section{Hello, World!}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

The \textsf{eqHello} redraw function renders six colored quads, rotating
around the origin. The \textsf{frameDraw} meth\-od provided by the
\textsf{eq::Channel} can be used as a convenience function to setup the
frustum and other OpenGL state. After setting up some lighting
parameters, \textsf{eqHello} rotates the scene and renders the quads
using immediate mode:

{\footnotesize\lstinputlisting[firstline=107, lastline=123]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

The \textsf{eqHello} main function sets up the communication with the
server, initializes and drives the rendering. The details of this setup
are explained in \sref{sEqPly}.

@@

{\footnotesize\lstinputlisting[firstline=57, lastline=66]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=55]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=70]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=71, lastline=78]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=81]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=83]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=61, lastline=74]
  {../../../Equalizer/examples/eqPly/initData.cpp}}

@@

The contextual information has to be used to render the view as
expected by Equalizer. Failure to use certain information will result in
incorrect rendering for some or all configurations. The channel render
context consist of:

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
\end{wrapfigure}
The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
are handled in the same way as window drawables. The event dispatch finds the
corresponding event handler for each received event, which is then used to
process the event in a fashion similar to AGL. The \textsf{GLXWindowEvent}
passes the \textsf{XEvent} to \textsf{GLXWindowIF::\-process\-Event}.

@@

{\footnotesize\lstinputlisting[firstline=124, lastline=140]
  {../../../Equalizer/libs/client/aglWindow.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=101]
  {../../../Equalizer/libs/client/glXWindow.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=178, lastline=224]
  {../../../Equalizer/libs/client/wglWindow.cpp}}

@@

\begin{wrapfigure}{r}{.612\textwidth}
  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
The render clients can access a slave instance of the view using
\textsf{Channel::getView}. When called from one of the frame task
methods, this method will return the view of the current destination
channel for which the task method is executed. Otherwise it returns the
channel's native view, if it has one. Only destination channels of an
active canvas have a native view.

Equalizer commits dirty layout entities at the beginning of each
\textsf{Config::startFrame}, and synchronizes the slave instances on the
render clients correctly with the current frame.

@@

The configuration contains stereo information in three places: the view, the
segment and the compound.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
  {\caption{\label{fStereoSwitch}Using two different decompositions during
      stereo and mono rendering}}
\end{wrapfigure}
The compound has an \textsf{eye} attribute defining which eyes it is capable of
updating. This allows to specify different compounds for the same channel,
depending on the stereo mode. One use case is to use one GPU each for updating a
stereo destination channel in stereo mode, and using a 2D compound with the two
GPUs in mono mode. \fig{fStereoSwitch} illustrates this example. The coloring in
mono mode is to illustrate the 2D decomposition. The default setting for the
compound eye attributes is all eyes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to a 2D (sort-first)
compound. The amount of pixel data for this part of the compositing
pipeline also approaches a constant value, i.e., the full frame buffer.

@@

Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes are blended during
rendering.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
Database decomposition has the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each channel),
fill rate (less samples per channel) and IO bandwidth for time-dependent data
(less data per time step and channel). Since the amount of texture memory needed
for each node decreases linearly, they make it possible to render data sets
which are not feasible to visualize with any other approach.

@@

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
Finding the correct assembly order for perspective frusta is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frusta. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8.3 for Equalizer 1.1.3+git \\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.8.1   & Mar 25, 2011 & Minor Revision\\
    1.8.2   & Apr 04, 2011 & Added \sref{sFocusDistance}\\
    1.8.3   & Aug 10, 2011 & Rewrote \sref{sHello}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\section{\label{sHello}Hello, World!}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the basic
principle of any Equalizer application: The application developer has to
implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
applications.

It can be run as a stand-alone application from the command line or using an
explicit configuration file for a visualization cluster. In the stand-alone
case, any Equalizer application including eqHello, will automatically
configure itself to use all graphics cards found on the local system for
scalability.

\subsection{Sequel}

Sequel is a thin layer on top of the canonical Equalizer programming
interface. It substantially simplifies the creation of parallel rendering
applications by implementing proven design patterns of Equalizer
applications. Sequel does not prohibit the usage of Equalizer or Collage
functionality, it simply makes it easier to use for the common use cases.

The main function instantiates an application object, initializes it, starts the
main loop and finally de-initializes the application:

{\footnotesize\lstinputlisting[firstline=58, lastline=66]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

\subsection{The Application}

The application object in Sequel represents one process in the cluster. The main
application instance has the rendering loop and controls all execution. All
other instances used for render client processes are passive and driven by
Equalizer.

The application is responsible for creating renderers. Sequel will request one
renderer for each GPU rendering thread. Sequel also provides automatic mapping
and synchronization of distributed objects, for which the application has to
provide a creation callback. Since \textsf{eqHello} does not need any
distributed objects it provides an empty implementation of
\textsf{createObject}:

{\footnotesize\lstinputlisting[firstline=47, lastline=55]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

\subsection{The Renderer}

The renderer is responsible for executing the application's rendering code. One
instance for each GPU is used. All calls to a single renderer are executed
serially and therefore thread-safe.

In \textsf{eqHello}, the renderer draws six colored quads. The only change from
a standard OpenGL application is the usage of the rendering context provided by
Equalizer, most notably the frustum and viewport. The rendering context is
described in detail in \sref{sRenderContext}, and \textsf{eqHello} simply calls
\textsf{applyRenderContext} which will execute the appropriate OpenGL calls.

After setting up lighting, the model is positioned using
\textsf{applyModelMatrix}. For convenience, Sequel maintains one camera per
view. The usage of this camera is purely optional, an application can implement
its own camera model.

The actual OpenGL code, rendering six colored quads, is omitted here for
brevity:

{\footnotesize\lstinputlisting[firstline=69, lastline=81]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=57, lastline=68]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=57]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=70, lastline=73]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=74, lastline=81]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=83, lastline=84]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=85]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=60, lastline=73]
  {../../../Equalizer/examples/eqPly/initData.cpp}}

@@

\label{sRenderContext}
The contextual information has to be used to render the view as
expected by Equalizer. Failure to use certain information will result in
incorrect rendering for some or all configurations. The channel render
context consist of:

@@

The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
\end{wrapfigure}
For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
are handled in the same way as window drawables. The event dispatch finds the
corresponding event handler for each received event, which is then used to
process the event in a fashion similar to AGL. The \textsf{GLXWindowEvent}
passes the \textsf{XEvent} to \textsf{GLXWindowIF::\-process\-Event}.

@@

{\footnotesize\lstinputlisting[firstline=138, lastline=154]
  {../../../Equalizer/libs/eq/client/agl/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=70, lastline=106]
  {../../../Equalizer/libs/eq/client/glx/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=180, lastline=226]
  {../../../Equalizer/libs/eq/client/wgl/window.cpp}}

@@

The render clients can access a slave instance of the view using
\textsf{Channel::getView}. When called from one of the frame task
methods, this method will return the view of the current destination
channel for which the task method is executed. Otherwise it returns the
channel's native view, if it has one. Only destination channels of an
active canvas have a native view.

\begin{wrapfigure}{r}{.612\textwidth}
  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
Equalizer commits dirty layout entities at the beginning of each
\textsf{Config::startFrame}, and synchronizes the slave instances on the
render clients correctly with the current frame.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
  {\caption{\label{fStereoSwitch}Using two different decompositions during
      stereo and mono rendering}}
\end{wrapfigure}
The configuration contains stereo information in three places: the view, the
segment and the compound.

@@

The compound has an \textsf{eye} attribute defining which eyes it is capable of
updating. This allows to specify different compounds for the same channel,
depending on the stereo mode. One use case is to use one GPU each for updating a
stereo destination channel in stereo mode, and using a 2D compound with the two
GPUs in mono mode. \fig{fStereoSwitch} illustrates this example. The coloring in
mono mode is to illustrate the 2D decomposition. The default setting for the
compound eye attributes is all eyes.

@@

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to a 2D (sort-first)
compound. The amount of pixel data for this part of the compositing
pipeline also approaches a constant value, i.e., the full frame buffer.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes are blended during
rendering.

Database decomposition has the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each channel),
fill rate (less samples per channel) and IO bandwidth for time-dependent data
(less data per time step and channel). Since the amount of texture memory needed
for each node decreases linearly, they make it possible to render data sets
which are not feasible to visualize with any other approach.

@@

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

Finding the correct assembly order for perspective frusta is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frusta. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.
